Motmot: A Distributed Chat Protocol

### 2012-05-04 paper/paper.tex
Oops #mergefail
.
More Paxos design

\usepackage{xspace}
\newcommand\Motmot{\textsc{Motmot}\xspace}

@@

Guaranteed delivery and total message ordering are perhaps the most fundamental
chat features provided by \Motmot.  In order to obtain these properties, the
clients participating in a chat must reach a consensus concerning the reception
and order of chat events.  \Motmot assumes a benign failure model and uses a
variant of the Paxos consensus protocol \cite{paxos}, which is the core of its
client library \texttt{libmotmot}, in order to provide these guarantees.

@@

@@

The \Motmot discovery server was implemented in Python using several open source
libraries. The server itself is a custom RPC-like reactor/dispatcher written
% TODO: cite gevent
using gevent, a coroutine-based networking library. For our communication
% TODO: cite msgpack
protocol, we use msgpack to serialize messages. This allows for easy
cross-language communication with clients and other servers, since
implementations are available for most popular languages.

@@

% TODO: cite pyopenssl
Finally, we used openssl (and the pyOpenSSL library in particular) to perform
certificate operations

@@

% TODO: EJ?
>>>>>>> Discovery server

@@

\item One cannot modify the main Pidgin event loop from within a plugin (without patching libpurple itself).  Instead, callbacks are made to plugin functions from within the main event loop, so finding a way to call libmotmot functions correctly (i.e. still honoring the correct control flow from libmotmot's perspective) and consistently within the bounds of these plugin functions was a second challenge.

@@

@@

@@

>>>>

\usepackage{url,xspace}
\newcommand\Motmot{\textsc{Motmot}\xspace}
\newcommand\libmotmot{\texttt{libmotmot}\xspace}

@@

Guaranteed delivery and total message ordering are perhaps the most fundamental
chat features provided by \Motmot.  In order to obtain these properties, the
clients participating in a chat must reach a consensus concerning the reception
and order of chat events.  \Motmot assumes a benign failure model and uses a
variant of the Paxos consensus protocol \cite{paxos,paxsimp}, which is the core
of its client library \libmotmot, in order to provide these guarantees.

@@

\libmotmot implements a slightly simplified variant of the Multi-Paxos
algorithm, augmented by a number of out-of-band yet nonetheless indispensible
side protocols.  This core Paxos implementation is wrapped by a client
interface and uses MessagePack \cite{msgpack} for network communication.
As with the other components of \Motmot, \libmotmot makes use of the GNOME
Library (GLib) \cite{glib}.

The Paxos implementation of \libmotmot makes a fundamental simplification to
the standard algorithm: we do not allow participants who leave the protocol, by
failure or otherwise, to rejoin.  In the original Paxos protocol, protocol
participants may fail benignly or depart the protocol at any time, and return
at some later point to continue voting and learning values in the system.
This is necessary in order for participants to maintain a notion of persistent
identity.  It is also needed to ensure recovery in the event of total system
failure.  Participants keep a persistent log of the values committed in the
protocol.  In the event of total failure, they can individually come back up,
rejoin the protocol, and corroborate on their individual copies of the log,
thereby allowing the protocol to continue where it left off.

A chat protocol, however, has no need for this mode of recovery, nor for this
particular notion of identity.  The persistent identity which users want from
a chat service can be maintained out-of-band.  If a user leaves a chat, there
is no need, at the level of the consensus protocol, to treat the user as the
\emph{same} user if they returns to the chat later.  Users need not be informed
of the messages that were exchanged in their absence.\footnote{We have not yet
investigated whether or not a distributed chat protocol is able to support
offline messages cleanly.}  So long as the system is able to determine, in a
consistent manner, that the user has left, the user can gain a new identity
within the protocol upon his or her return.

Similarly, if a chat session ends, either due to failure or simply because all
users leave, the termination is allowed to be final.  The participants in the
chat may start a new session if they wish, but this new session need not have
any memory nor knowledge of the old one.

This simplification to Paxos yields many benefits.  For instance, clients may
still choose to keep a record of chats in a local file, but by disallowing
protocol rejoins, this persistent storage does not play any role in the safety
or progress of \libmotmot's Paxos implementation.  We can also avoid the need
for complicated and high-bandwidth recovery mechanisms.

\libmotmot also implements a number of out-of-band protocols which support
the core Paxos algorithm.  We will now describe each of these side protocols.

@@

The \Motmot discovery server was implemented in Python using several open source
libraries. The server itself is a custom RPC-like reactor/dispatcher written
using GEvent \cite{glib}, a coroutine-based networking library. For our
communication protocol, we use MessagePack \cite{msgpack} to serialize messages.
This allows for easy cross-language communication with clients and other servers,
since implementations are available for most popular languages.

@@

% TODO: cite pyopenssl
Finally, we used openssl (and the pyOpenSSL library in particular) to perform
certificate operations.

@@

% TODO: EJ?

@@

\item The main Pidgin event loop cannot be modified from within a plugin (without patching libpurple itself).  Instead, callbacks are made to plugin functions from within the main event loop, so we had to find a way to call libmotmot functions correctly (i.e. still honoring the correct control flow from libmotmot's perspective) and consistently within the bounds of these plugin functions

\item libmotmot and libpurple use different handles to differentiate between chats and users, and translating between the two handles proved challenging, especially given the complexity of the information associated with libpurple chats (e.g. a single chat may have an id, a room name, PurpleConnection and PurpleConvChat objects...)

@@

The bulk of the integration work was writing callbacks for the motmot\_init() function of libmotmot.  Once these callbacks are written correctly, libmotmot handles most of the work of starting, sending, and coordinating chats automatically.  These callbacks were, in almost all instances, responsible for calling libpurple functions, which handle the UI aspect of the associated libmotmot calls.  In order to pass relevant information to our callbacks, we created a MotmotInfo struct, a large data struct that is used differently in different callbacks to get all the necessary info to the libpurple calls.

We faced a challenge when writing the enter() function.  This function is a callback responsible for allowing a user to enter a chat.  Due to the structure of libmotmot, the only information passed into enter() as an argument is an internal handle, and we need create some sort of external handle.  We solved this by creating a global variable, GLOBAL\_ACCOUNT, which represents the account that a user first logs in under, and we associate our external handle with this GLOBAL\_ACCOUNT.  Though this ``works'' if a user only signs in with on eaccount, this breaks when a user decides to sign in with multiple accounts at once (something that libpurple allows), and should be addressed in future versions.

@@

  \bibitem{glib}
    GNOME Library.
    \url{http://developer.gnome.org/about/}, 2011.

@@

  \bibitem{msgpack}
    MessagePack.
    \url{http://www.msgpack.org/}, 2010.
